On the uniqueness and stability of dictionaries \\ for sparse representation of noisy signals

### 2019-05-10 manuscripts/IEEE/final_submission.tex
sml and fig

\begin{abstract}
%Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, there are few universal conditions guaranteeing the stability of such features in the presence of noise. Here, we prove very generally that optimal dictionaries and sparse codes are unique up to measurement error for all diverse enough datasets generated by the sparse coding model. Applications are given to data analysis, engineering, and neuroscience. 
Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, universal guarantees of the stability of such features in the presence of noise are lacking. Here, we provide very general conditions guaranteeing when dictionaries yielding the sparsest encodings are unique and stable with respect to measurement or modeling error. We demonstrate that some or all dictionary elements are recoverable from noisy data even if the dictionary fails to satisfy the spark condition, its size is overestimated, or it induces only polynomial number of distinct sparse supports for the data. Importantly, we derive these guarantees without requiring any constraints on the recovered dictionary beyond a natural upper bound on its size. Our results yield an effective procedure sufficient to affirm if a proposed solution to the dictionary learning problem is unique within bounds commensurate with the noise. We suggest applications to data analysis, engineering, and neuroscience and close with some remaining challenges left open by our work.
\end{abstract}

@@

Note that every solution to Prob.~\ref{InverseProblem} represents infinitely many equivalent alternatives $\mathbf{BPD}$ and $\mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_N$ parametrized by a choice of permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$. Identifying these ambiguities (labelling and scale) yields a set of elementary waveforms (the columns of $\mathbf{B}$) and their associated sparse coefficients (the entries of $\mathbf{\overline x}_i$) for approximating each data point $\mathbf{z}_i$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=.24 \linewidth]{figs/basis.png}
% \includegraphics[width=.25 \linewidth]{figs/Ahat_FastICA_M640_K8_noise0_2042}
\includegraphics[width=.24 \linewidth]{figs/Ahat_FastICA_M640_K8_noise0_4083.png}
\includegraphics[width=.24 \linewidth]{figs/Ahat_FastICA_M640_K8_noise1_0205.png}
\includegraphics[width=.24 \linewidth]{figs/Ahat_FastICA_M640_K8_noise2_0409.png}
\caption{\textbf{Learning a dictionary from increasingly noisy data}. The (unraveled) basis elements of the $8 \times 8$ discrete cosine transform (DCT) form the 64 columns of the left-most matrix above. Three increasingly imprecise dictionaries (columns reordered to best match original) are recovered by FastICA \cite{hyvarinen2000independent} trained on data generated from $8$-sparse linear combinations of DCT elements corrupted with additive noise (increasing from left to right).}
\vspace{-.6 cm}
\label{noisyrecovery}
\end{center}
\end{figure}

>>>>

\begin{abstract}
%Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, there are few universal conditions guaranteeing the stability of such features in the presence of noise. Here, we prove very generally that optimal dictionaries and sparse codes are unique up to measurement error for all diverse enough datasets generated by the sparse coding model. Applications are given to data analysis, engineering, and neuroscience. 
Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, universal guarantees of the stability of such features in the presence of noise are lacking. Here, we provide very general conditions guaranteeing when dictionaries yielding the sparsest encodings are unique and stable with respect to measurement or modeling error. We demonstrate that some or all dictionary elements are recoverable from noisy data even if the dictionary fails to satisfy the spark condition, its size is overestimated, or only a polynomial number of distinct sparse supports appear in the data. Importantly, we derive these guarantees without requiring any constraints on the recovered dictionary beyond a natural upper bound on its size. Our results yield an effective procedure sufficient to affirm if a proposed solution to the dictionary learning problem is unique within bounds commensurate with the noise. We suggest applications to data analysis, engineering, and neuroscience and close with some remaining challenges left open by our work.
\end{abstract}

@@

Note that every solution to Prob.~\ref{InverseProblem} represents infinitely many equivalent alternatives $\mathbf{BPD}$ and $\mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_N$ parametrized by a choice of permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$. 
Identifying these ambiguities (labelling and scale) yields a single orbit of solutions represented by any particular set of elementary waveforms (the columns of $\mathbf{B}$) and their associated sparse coefficients (the entries of $\mathbf{\overline x}_i$) that reconstruct each data point $\mathbf{z}_i$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=.24 \linewidth]{figs/basis.png}
% \includegraphics[width=.25 \linewidth]{figs/Ahat_FastICA_M640_K8_noise0_2042}
%\includegraphics[width=.24 \linewidth]{figs/Ahat_FastICA_M640_K8_noise0_4083.png}
\includegraphics[width=.24 \linewidth]{figs/Ahat_FastICA_M640_K8_noise1_0205.png}
\includegraphics[width=.24 \linewidth]{figs/Ahat_FastICA_M640_K8_noise2_0409.png}
\includegraphics[width=.24 \linewidth]{figs/Ahat_FastICA_M640_K8_noise5_1021.png}
\caption{\textbf{Learning a dictionary from increasingly noisy data}. The (unraveled) basis elements of the $8 \times 8$ discrete cosine transform (DCT) form the 64 columns of the left-most matrix above. Three increasingly imprecise dictionaries (columns reordered to best match original) are recovered by FastICA \cite{hyvarinen2000independent} trained on data generated from $8$-sparse linear combinations of DCT elements corrupted with additive noise (increasing from left to right).}
\vspace{-.6 cm}
\label{noisyrecovery}
\end{center}
\end{figure}
